---
title: "Exploratory data analysis"
author: "Ivan Prodaiko"
subtitle: Data visualization hw3
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
```{r}
loadLibs <- function() {
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggbeeswarm)
  library(ggridges)
  library(ggthemes)
  library(ggrepel)
  library(reshape2)
  library(purrr)
  library(Hmisc)
}
suppressMessages(loadLibs())

remove_outliers <- function(x, na.rm = TRUE, ...) {
  qnt <- quantile(x, probs=c(.05, .95), na.rm = na.rm, ...)
  H <- 1.5 * IQR(x, na.rm = na.rm)
  y <- x
  y[x < (qnt[1] - H)] <- NA
  y[x > (qnt[2] + H)] <- NA
  y
}
```
### Load dataset:
```{r}
df <- read.csv("./music.csv", header = T)
head(df)
```
### Lets first of all look the distributions of main features:
```{r}
mainfeatures <- df %>% select(c("artist.hotttnesss", "duration", "tempo", "song.hotttnesss", "loudness", "familiarity"))

as.data.frame(sapply(mainfeatures, remove_outliers)) %>% 
  filter_at(vars(song.hotttnesss), any_vars(. != 0)) %>%
  na.omit() %>%
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_density() +
  theme_fivethirtyeight() +
  theme(
    text = element_text(family = 'Ubuntu', size = 12)
  ) +
  labs(
    title = "Destribution of main features",
    caption = "Million songs dataset"
  )
```

Several of tham are distributed almost normaly while some of tham are very skewed.

### Lets check what mode is dominant in the dataset and how the mode slection impacts the popularity of the songs:
```{r}
df$modeString[df$mode==1] <- "Major"
df$modeString[df$mode==0] <- "Minor"

ggplot(df, aes(x=modeString)) +
  geom_histogram(stat = "count", fill = "dark blue", alpha=0.4)+
  theme_fivethirtyeight() +
  theme(text = element_text(family = 'Ubuntu', size = 12)) +
  labs(
    title = "Modes count",
    caption = "Million songs dataset"
   )

df %>% 
  select(modeString, song.hotttnesss) %>% 
  filter_at(vars(song.hotttnesss), any_vars(. != 0)) %>%
  na.omit() %>% 
  ggplot(aes(x=modeString, y=song.hotttnesss)) + 
  geom_violin(alpha=.1, color='dark blue') + 
  geom_jitter(shape=16, position=position_jitter(0.3), alpha = 0.15) +
  geom_boxplot(width=0.1, color='dark red') +
  coord_flip() +
  theme_fivethirtyeight() +
  theme(
    text = element_text(family = 'Ubuntu', size = 12)
  ) +
  labs(
    title = "Destribution of song hotness related to its mode",
    caption = "Million songs dataset"
  )
```

As we can conclude, we songs in **minor** is little bit **more popular**, than songs is **major**. But overall pattern is alomost identical.

### Lets see how median song duration changed over time:
```{r}
dgr_fmt <- function(x) {
  parse(text = sprintf("%.2f~minutes", x/60))
}

df2 <- df[df$year != 0,] %>% select(c("duration", "year"))

aggregate(df2, by=list(df2$year), median) %>%
  ggplot(aes(x=year)) +
    geom_point(mapping = aes(y=duration), alpha = 0.8) +
    geom_smooth(method = "gam", formula = y ~ s(x), mapping = aes(y=duration), se=F, color='darkred') +
    theme_fivethirtyeight() +
    theme(text = element_text(family = 'Ubuntu', size = 12)) +
    scale_x_continuous(
      breaks = seq(min(df2$year), max(df2$year), 10),
      expand = c(0, 0, 0, 0),
      minor_breaks = F
    ) +
    scale_y_continuous(
      label = dgr_fmt
    ) +
    geom_curve(data = data.frame(x = 1986, y = 3 * 60, xend = 1965 + 1, yend = median(df2[df2$year == 1965,]$duration)),
            aes(x = x, y = y, xend = xend, yend = yend),
            curvature = -0.1,
            arrow = arrow(angle = 15, length = unit(5, 'pt')),
            inherit.aes = FALSE) +
    geom_text(
      data = data.frame(x = 1980, y = 3 * 60, label = '"The Beatles" rise\nRock-n-roll revolution'),
      aes(x = x, y = y, label = label),
      inherit.aes = FALSE, nudge_x = 16
    ) +
    geom_curve(data = data.frame(x = 1960, y = 4.17 * 60, xend = 1970, yend = median(df2[df2$year == 1970,]$duration)),
            aes(x = x, y = y, xend = xend, yend = yend),
            curvature = -0.1,
            arrow = arrow(angle = 20, length = unit(5, 'pt')),
            inherit.aes = FALSE) +
    geom_text(
      data = data.frame(x = 1960, y = 4.17 * 60, label = 'First "Black Subbuth" albom\nHeavy metal era started'),
      aes(x = x, y = y, label = label),
      inherit.aes = FALSE, nudge_x = -12
    ) +
    labs(
      title = "Avarage song duration",
      subtitle = "Changing over time",
      caption = "Million songs dataset"
    )
```

As we could see avarage duration of songs has songs tend to be **below 3 minutes** before **mid 60th**. After the rock-n-roll rise it become **approximately 4 minutes since than**.

### Lets see how median song tempo changed over time:
```{r}
dgr_fmt <- function(x) {
  parse(text = sprintf("%.2f~BPM", x))
}

df2 <- df[df$year != 0,] %>% select(c("tempo", "year"))

aggregate(df2, by=list(df2$year), median) %>%
  ggplot(aes(x=year)) +
    geom_point(mapping = aes(y=tempo), alpha = 0.8) +
    geom_smooth(method = "gam", formula = y ~ s(x), mapping = aes(y=tempo), se=F, color='darkred') +
    theme_fivethirtyeight() +
    theme(text = element_text(family = 'Ubuntu', size = 12)) +
    scale_x_continuous(
      breaks = seq(min(df2$year), max(df2$year), 10),
      expand = c(0, 0, 0, 0),
      minor_breaks = F
    ) +
    scale_y_continuous(
      label = dgr_fmt
    ) +
    labs(
      title = "Avarage song tempo",
      subtitle = "Changing over time",
      caption = "Million songs dataset"
    )
```

Avarage tempo of songs has **started raising** from **mid 40th** and reached **120 BPM** and approximately stays on this level now.

### Lets see how populer different genres of music are:
```{r}
genres <- as_data_frame(table(genre = df[df$terms != 'Not available',]$terms))
genres[order(-genres$n), ] %>% slice(1:30) %>%
  ggplot(aes(x=reorder(genre, n), y=n)) +
  geom_bar(stat = "identity", fill = "dark blue", alpha=0.4)+
  coord_flip()+
  theme_fivethirtyeight() +
  theme(text = element_text(family = 'Ubuntu', size = 12)) +
  scale_y_continuous(breaks = NULL) +
  labs(
    title = "Most popular genres",
    caption = "Million songs dataset"
   )
```

As we could see **rock** and **hip-hop** are the most popular.

### Lets where the artists are from:
```{r}
geo_available <- as_data_frame(table(location = df[df$location != 'Not available',]$location))
geo_available <- geo_available[order(-geo_available$n), ]

most_popular_location <-
  geo_available[grep(",", geo_available$location),] %>% 
  slice(1:30)

most_popular_location %>% ggplot(aes(x=reorder(location, n), y=n)) +
  geom_bar(stat = "identity", fill = "dark blue", alpha=0.4) +
  coord_flip()+
  theme_fivethirtyeight() +
  theme(text = element_text(family = 'Ubuntu', size = 12)) +
  scale_y_continuous(breaks = NULL) +
  labs(
    title = "Most popular locations",
    caption = "Million songs dataset"
   )
```

As we could see **UK** and **USA** creates the most popular music (according to this dataset)

### Lets draw the distribution of songs over the worlds map:
```{r}
df3 <- df[df$location != 'Not available',]
WorldData <- map_data('world')
WorldData %>% filter(region != "Antarctica") -> WorldData
WorldData <- fortify(WorldData)

ggplot() + 
  geom_map(data=WorldData, map=WorldData,
    aes(x=long, y=lat, group=group, map_id=region),
    fill="white", colour="#7f7f7f", size=0.5
  ) +
  geom_point(data = df3, aes(x = longitude, y = latitude), col="dark red", size=.5, alpha = .8) +
  theme_fivethirtyeight() +
  theme(text = element_text(family = 'Ubuntu', size = 12)) +
  labs(
    title = "Concentration of artists around on the worlds map",
    caption = "Million songs dataset"
   )
```

Most of the music are produced in **USA east coast** and **Western Europe** (according to dataset).

